Types from data: \textnormal{Making structured data first-class citizens in F\#}

### 2015-11-20 pldi-2016/paper.tex

In this paper, we integrate external structured data into the F\# type system. As most real-world data
does not come with an explicit schema, we develop a shape inference algorithm that infers a shape from 
representative sample documents and integrate it into the F\# type system using type providers.

@@

Social network clients, applications for finding tomorrow's weather or searching train schedules
all communicate with external services. Increasing number of such services provide REST-based 
end-points that return data as CSV, XML or JSON. Despite many schematization efforts, most 
services do not come with an explicit schema. At best, the documentation provides sample responses 
for typical requests.

@@

Using the JSON type provider from F\# Data, we can write code with exactly the 
same functionality in two lines:
%
\vspace{-0.1em}
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{W} = \ident{JsonProvider}\langl\str{http://api.owm.org/?q=NYC}\rangl \\[0.1em]
 \ident{printfn}~\str{Lovely \%f!}~(\ident{W.GetSample().Main.Temp})
\end{array}
\end{equation*}
%
$\ident{JsonProvider}\langl\str{...}\rangl$ invokes a type provider at 
compile-time with the URL as a sample. The type provider infers the structure of the response
and provides a type with a \ident{GetSample} method that returns a parsed JSON with nested
properties \ident{Main.Temp}, returning the temperature as a number.

In short, \emph{the types come from the sample data}. In our experience, this technique is 
immensely practical and surprisingly effective in achieving sound information interchange 
in heterogeneous systems. This paper presents our approach:

@@

\item We give a formal model (\S\ref{sec:formal}) and use it to prove
  \emph{relativized type safety} for the type providers (\S\ref{sec:safety}).
  This adapts the ML-style type safety for the context of the web.
\end{itemize}

@@

\item The type providers prefer F\# objects with properties. This is to allow extensible 
  (open-world) data formats (\S\ref{sec:providers-xml}). It also interacts well with developer tooling
  as most F\# editors provide auto-completion on ``.'' and so objects are easier to use than 
  types that require pattern matching.

\item Finally, the mechanism handles practical concerns important in the real world. This includes 
  support for different numerical types, \kvd{null} values and missing data.
\end{itemize}

@@

The JSON format is a popular data exchange format based on 
JavaScript data structures. The following is the definition of \ident{JsonValue} 
used earlier (\S\ref{sec:introduction}) to represent JSON data:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{JsonValue} = \\[0.1em]
 \quad|~ \ident{Number}~\kvd{of}~\ident{float} ~~~~|~ \ident{Boolean}~\kvd{of}~\ident{bool} \\[0.1em]
 \quad|~ \ident{String}~\kvd{of}~\ident{string} ~~\qquad\, |~ \ident{Null} \\[0.1em]
 \quad|~ \ident{Record}~\kvd{of}~\ident{Map}\langl\ident{string}, \ident{JsonValue}\rangl \\[0.1em]
 \quad|~ \ident{Array}~\kvd{of}~\ident{JsonValue}[] \\[0.1em]
\end{array}
\end{equation*}
%
The earlier example used only a nested record containing a number. To demonstrate other 
aspects of the JSON type provider, we look at an example that also involves an array:
%
{\small{
\begin{verbatim}
  [ { "name":"Jan", "age":25 }, { "name":"Tomas" },
    { "name":"Alexander", "age":3.5 } ]
\end{verbatim}
}}
%
\noindent
The standard approach to print the names and ages would be to pattern match on the parsed 
\ident{JsonValue}, check that the top-level node is a \ident{Array} and iterate over the elements 
checking that each element is a \ident{Record} with certain properties. We would throw an exception 
for values of an incorrect shape. As before, the code would specify field names as strings, which 
is error prone and can not be statically checked.

@@

\paragraph{Type providers.}
The notation $\ident{JsonProvider}\langl\str{people.json}\rangl$ passes a \emph{static parameter} 
to the type provider. Static parameters are resolved at compile-time and have to be constant. 
The provider analyzes the sample and generates a type  \ident{People}. Most F\# editors also 
execute the type provider in the background at development-time and use the provided types 
in auto-completion and background type-checking.

@@

Consider a simple extensible document format where a root element {\ttfamily\small <doc>} can 
contain a number of document elements, one of which is {\ttfamily\small <heading>} representing 
headings:
%
{\small{
\begin{verbatim}
  <doc>
    <heading>Working with JSON</heading>
    <p>Type providers make this easy.</p>
    <heading>Working with XML</heading>
    <p>Processing XML is as easy as JSON.</p>
    <image source="xml.png" />
  </doc>
\end{verbatim}
}}
%
\noindent
The F\# Data library has been designed primarily to simplify reading of data. For example,
say we want to print all headings in the document. The sample shows a part of the document structure 
(in particular the {\ttfamily\small <heading>} element), but it does not show all possible elements 
(say, {\ttfamily\small <table>}). Assuming the above document is \strf{sample.xml}, we can write:
%
\noindent
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Document}~=~\ident{XmlProvider}\langl\str{sample.xml}\rangl\hspace{1em} \\[0.5em]
 \kvd{let}~\ident{root}~=~\ident{Document.Load}(\str{c:/pldi/another.xml})\\
 \kvd{for}~\ident{elem}~\kvd{in}~\ident{root.Doc}~\kvd{do}\\
 \quad\ident{Option.iter}~(\ident{printf}~\str{ - \%s})~\ident{elem.Heading}\\
\end{array}
\end{equation*}
%
The example iterates over a collection of elements returned by \ident{root.Doc}. The type of \ident{elem} 
provides a typed access to elements known from the sample and so we can write \ident{elem.Heading}, 
which returns an optional string value.

@@

\subsection{Summary}
\label{sec:providers-sum}
Throughout the introduction, we used data sets that demonstrate the typical problems that are frequent
in the real-world (\emph{missing data}, \emph{inconsistent encoding} of primitive values and 
\emph{heterogeneous shapes}). In our experience, these are the most common issues. The following 
JSON response with government debt information returned by the World Bank\footnote{Available at 
\url{http://data.worldbank.org}} demonstrates all three problems:
%
{\small{
\begin{verbatim}
  [ { "page": 1, "pages": 5 },
    [ { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2012", "value": null },
      { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2010", "value": "35.14229" } ] ]
\end{verbatim}
}}
%
\noindent
First of all, the top-level element is a collection containing two values of different kind.
The first is a record with meta-data about the current page and the second is an array with data. 
The actual F\# Data implementation supports a concept of heterogeneous collections (briefly outlined
in \S\ref{sec:impl-hetero}) and provides a type with properties \ident{Record} for the former and 
\ident{Array} for the latter. Second, the \ident{value} field is \kvd{null} for some records. Third, 
numbers can be represented in JSON as numeric literals (without quotes), but here, they are 
returned as string literals instead.\footnote{This is often used to avoid non-standard numerical 
types of JavaScript.}

In addition to type providers for JSON and XM, F\# Data also implements a type provider for CSV 
(\S\ref{sec:impl-parsing}). We treat CSV files as lists of records (with field for each column) 
and so CSV is handled directly by our inference algorithm.

@@

@@

\noindent
The rule that allows preferred shapes to have fewer fields (\ref{eq:sub-record3}) is particularly
important. This is important, because it allows us to prefer records over the \kvd{any} 
shape. Given $\{ \ident{name}\!:\!\ident{string} \}$ and $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int} \}$,
we find a common shape $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int}~\kvd{option} \}$.
This is a least upper bound and more usable than the top shape $\kvd{var}$.

Some of the aspects of our system (numeric types and handling of missing data) offer a range of
choices. For example, OCaml also has different numerical types, but would likely always handle 
\kvd{null} explicitly.

@@

\begin{theorem}[Preferred shape function]
\label{thm:func}
For all $\sigma_1$ and $\sigma_2$ there exists exactly one $\sigma$ such that 
$\sigma_1 \tsep \sigma_2 \vdash \sigma$. 
\end{theorem}
\begin{proof}
The pre-conditions of rules in Figure~\ref{fig:subtyping-cst} are disjoint, with the exception 
of (\emph{sym}) and (\emph{refl}), but applying those does not affect the result. Reflexivity is
preserved recursively and all rules, symmetry does not enable different derrivations, even for
non-symmetric (\emph{nullable}) and (\emph{prim}).
\end{proof}

>>>>

In this paper, we integrate external structured data into F\# programming. As most real-world data
does not come with an explicit schema, we develop a shape inference algorithm that infers a shape from 
representative sample documents and transforms the shape into F\# types using type providers.
We present a formalization and a relative type soundness theorem for a simplified version of this process.

@@

Applications for social network clients, finding tomorrow's weather or searching train schedules
all communicate with external services. Increasingly these services provide REST-based 
end-points that return data as CSV, XML or JSON. Despite many schematization efforts, most 
services do not come with an explicit schema. At best, the documentation provides sample responses 
for typical requests.

@@

Using the JSON type provider from F\# Data, we can write code with exactly the 
same functionality in two lines:
%
\vspace{-0.1em}
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{W} = \ident{JsonProvider}\langl\str{http://api.owm.org/?q=NYC}\rangl \\[0.1em]
 \ident{printfn}~\str{Lovely \%f!}~(\ident{W.GetSample().Main.Temp})
\end{array}
\end{equation*}
%
$\ident{JsonProvider}\langl\str{...}\rangl$ invokes a type provider \cite{fsharp-typeprov} at 
compile-time with the URL as a sample. The type provider infers the structure of the response
and provides a type with a \ident{GetSample} method that returns a parsed JSON with nested
properties \ident{Main.Temp}, returning the temperature as a number.

In short, \emph{the types come from the sample data}. In our experience, this technique is 
both practical and surprisingly effective in achieving more sound information interchange 
in heterogeneous systems. Our contributions are as follows:

@@

\item We give a formal model (\S\ref{sec:formal}) and use it to prove
  \emph{relativized type safety} for the type providers (\S\ref{sec:safety}).
\end{itemize}

@@

\item The type providers prefer F\# object types with properties. This allows extensible 
  (open-world) data formats (\S\ref{sec:providers-xml}). It also interacts well with developer tooling,
  as most F\# editors provide auto-completion on ``.'' making values of provided types easy to use.
  This also makes our techniques applicable any language with nominal object types -- for example
  variations of Java or C# with a type provider mechanism added.

\item Finally, the mechanism handles practical concerns including
  support for different numerical types, \kvd{null} values and missing data.
\end{itemize}

@@

The JSON format is a popular data exchange format based on 
JavaScript data structures. The following is the definition of \ident{JsonValue} 
used earlier (\S\ref{sec:introduction}) to represent JSON data:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{JsonValue} = \\[0.1em]
 \quad|~ \ident{Number}~\kvd{of}~\ident{float} ~~~~|~ \ident{Boolean}~\kvd{of}~\ident{bool} \\[0.1em]
 \quad|~ \ident{String}~\kvd{of}~\ident{string} ~~~~~~ |~ \ident{Null} \\[0.1em]
 \quad|~ \ident{Record}~\kvd{of}~\ident{Map}\langl\ident{string}, \ident{JsonValue}\rangl \\[0.1em]
 \quad|~ \ident{Array}~\kvd{of}~\ident{JsonValue}[] \\[0.1em]
\end{array}
\end{equation*}
%
The earlier example used only a nested record containing a number. To demonstrate other 
aspects of the JSON type provider, we look at an example that also involves an array:
%
{\small{
\begin{verbatim}
  [ { "name":"Jan", "age":25 }, { "name":"Tomas" },
    { "name":"Alexander", "age":3.5 } ]
\end{verbatim}
}}
%
\noindent
The standard approach to print the names and ages would be to pattern match on the parsed 
\ident{JsonValue}, check that the top-level node is a \ident{Array} and iterate over the elements 
checking that each element is a \ident{Record} with certain properties. We would throw an exception 
for values of an incorrect shape. As before, the code would specify field names as strings, which 
is error prone and can not be statically checked.

@@

\paragraph{Type providers.}
The notation $\ident{JsonProvider}\langl\str{people.json}\rangl$ passes a \emph{static parameter} 
to the type provider. Static parameters are resolved at compile-time and have to be constant. 
The provider analyzes the sample and provides a type  \ident{People}. F\# editors also 
execute the type provider in the background at development-time and use the provided types 
in auto-completion and background type-checking.

@@

Consider a simple extensible document format where a root element {\ttfamily\small <doc>} can 
contain a number of document elements, one of which is {\ttfamily\small <heading>} representing 
headings:
%
{\small{
\begin{verbatim}
  <doc>
    <heading>Working with JSON</heading>
    <p>Type providers make this easy.</p>
    <heading>Working with XML</heading>
    <p>Processing XML is as easy as JSON.</p>
    <image source="xml.png" />
  </doc>
\end{verbatim}
}}
%
\noindent
The F\# Data library has been designed primarily to simplify reading of data. For example,
say we want to print all headings in the document. The sample shows a part of the document structure 
(in particular the {\ttfamily\small <heading>} element), but it does not show all possible elements 
(say, {\ttfamily\small <table>}). Assuming the above document is \strf{sample.xml}, we can write:
%
\noindent
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Document}~=~\ident{XmlProvider}\langl\str{sample.xml}\rangl\hspace{1em} \\[0.5em]
 \kvd{let}~\ident{root}~=~\ident{Document.Load}(\str{c:/pldi/another.xml})\\
 \kvd{for}~\ident{elem}~\kvd{in}~\ident{root.Doc}~\kvd{do}\\
 \quad\ident{Option.iter}~(\ident{printf}~\str{ - \%s})~\ident{elem.Heading}\\
\end{array}
\end{equation*}
%
The example iterates over a collection of elements returned by \ident{root.Doc}. The type of \ident{elem} 
provides typed access to elements known from the sample and so we can write \ident{elem.Heading}, 
which returns an optional string value.

@@

\subsection{Summary}
\label{sec:providers-sum}
Throughout the introduction, we used data sets that demonstrate the typical problems that are frequent
in the real-world (\emph{missing data}, \emph{inconsistent encoding} of primitive values and 
\emph{heterogeneous shapes}). The following 
JSON response with government debt information returned by the World Bank\footnote{Available at 
\url{http://data.worldbank.org}} demonstrates all three problems:
%
{\small{
\begin{verbatim}
  [ { "page": 1, "pages": 5 },
    [ { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2012", "value": null },
      { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2010", "value": "35.14229" } ] ]
\end{verbatim}
}}
%
\noindent
First of all, the top-level element is a collection containing two values of different kind.
The first is a record with meta-data about the current page and the second is an array with data. 
The actual F\# Data implementation supports a concept of heterogeneous collections (briefly outlined
in \S\ref{sec:impl-hetero}) and provides a type with properties \ident{Record} for the former and 
\ident{Array} for the latter. Second, the \ident{value} field is \kvd{null} for some records. Third, 
numbers can be represented in JSON as numeric literals (without quotes), but here, they are 
returned as string literals instead.\footnote{This is often used to avoid non-standard numerical 
types of JavaScript.}

In addition to type providers for JSON and XML, F\# Data also implements a type provider for CSV 
(\S\ref{sec:impl-parsing}). We treat CSV files as lists of records (with field for each column) 
and so CSV is handled directly by our inference algorithm.

@@

The shape algebra and inference presented here is influenced by the design principles
we outlined earlier and the type definitions available in the F\# language.
If applied to other languages, details may differe, for example
w.r.t. numerical types and missing data.

@@

\noindent
The rule that allows preferred shapes to have fewer fields (\ref{eq:sub-record3}) is particularly
important. This allows us to prefer records with optional fields over the \kvd{any} 
shape. Given $\{ \ident{name}\!:\!\ident{string} \}$ and $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int} \}$,
we find a common shape $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int}~\kvd{option} \}$.
This is a least upper bound and more usable than the top shape $\kvd{var}$.

@@

\begin{theorem}[Preferred shape function]
\label{thm:func}
For all $\sigma_1$ and $\sigma_2$ there exists exactly one $\sigma$ such that 
$\sigma_1 \tsep \sigma_2 \vdash \sigma$. 
\end{theorem}
\begin{proof}
The pre-conditions of rules in Figure~\ref{fig:subtyping-cst} are disjoint, with the exception 
of (\emph{sym}) and (\emph{refl}), but applying those does not affect the result. Reflexivity is
preserved recursively and all rules, symmetry does not enable different derivations, even for
non-symmetric (\emph{nullable}) and (\emph{prim}).
\end{proof}
